---
title: "RNA-Seq Analysis Project"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center"
)
```

# 0. Introduction

## 0.1 Project Overview

This RNA-Seq analysis looks at gene expression patterns in human tissues, focusing on finding genes that are uniquely or highly expressed in specific individuals.

## 0.2 Dataset Description

The dataset contains RNA-Seq data from human tissues:
- Gene expression counts (28,391 genes × 18 samples) 
- Multiple tissue types: liver, brain, muscle, heart, kidney, lung
- Three donors: S7, S12, S13

### References

-   Zambelli et al 2022 (aScan: A Novel Method for the Study of Allele Specific Expression in Single Individuals)
-   Picardi et al 2015 (Profiling RNA editing in human tissues: towards the inosinome Atlas)
-   D'Erchia et al 2014 (Tissue-specific mtDNA abundance from exome data and its correlation with mitochondrial transcription, mass and respiratory activity)

# 1. Data Processing and Quality Control

## 1.1 Initial Setup

Loading the required packages for the analysis.

- **edgeR**: For differential expression analysis and TMM normalization
- **ggplot2**: For creating plots and visualizations
- **tidyverse**: For data manipulation and processing
- **stringr**: For handling gene names and text processing

```{r libraries}
# Load required libraries
library(edgeR)
library(ggplot2)
library(tidyverse)
library(stringr)

# Create results directory if it doesn't exist
dir.create("../results", showWarnings = TRUE)
```

## 1.2 Data Import and Preprocessing

Loading the data and checking it looks correct.

### Load Data
```{r load_data}
# Read metadata and counts
metadata <- read.csv("../data/Metadata2025.csv", sep = "\t", header = TRUE, check.names = FALSE, row.names = 1)
counts <- read.csv("../data/Counts2025.csv", sep = "\t", header = TRUE, check.names = FALSE, row.names = 1)
```

### Control Data
```{r control_data}
head(counts) # First few rows
dim(counts) # Should be 28391 genes x 18 samples
apply(counts, 2, sum) # Total counts per sample
sum(is.na(counts)) # Check for missing values
head(metadata) # First few rows
dim(metadata) # Should be 18 samples x 2 columns
sum(is.na(metadata)) # Check for missing values
```

### Add Sample Column to Metadata
Making sure sample names match between metadata and count data.
```{r add_sample_column_to_metadata}
metadata$Sample <- paste(metadata$Tissue, metadata$Donor, sep = "_")

columns_name_counts_match_check <- identical(metadata$Sample, colnames(counts))

if (columns_name_counts_match_check == FALSE) {
    metadata <- metadata[match(colnames(counts), metadata$Sample), ]
    rownames(metadata) <- NULL
    print("Reordered metadata to match count data.")
} else {
    print("Sample names already match.")
}

head(metadata)

columns_name_counts_match_check <- identical(metadata$Sample, colnames(counts))

if (columns_name_counts_match_check == FALSE) {
    stop("Sample names still don't match!")
} else {
    print("All good - sample names match.")
}
```

### Create a DGEList object for edgeR analysis
Setting up the main data object for analysis.

```{r create_dge}
dge <- DGEList(counts = counts, samples = metadata)
head(dge$counts) # Count matrix
head(dge$samples) # Sample info
```

# 2. Filtering
Removing genes with very low expression to reduce noise.
```{r filtering}
# Keep genes with at least 10 counts in at least one sample
keep <- rowSums(dge$counts > 10) >= 1
dge <- dge[keep, , keep.lib.sizes = FALSE]
print("Dimensions after filtering:")
dim(dge)
```

# 3. Normalization
Normalizing for library size differences using TMM method.
```{r normalizing}
dge <- calcNormFactors(dge)

head(dge$samples)
```

# 4. Multi-dimensional Scaling
Looking at how similar samples are to each other.

```{r mds_analysis}
mds <- plotMDS(dge, plot = FALSE)

# Plot with ggplot2
ggplot(data.frame(
    x = mds$x,
    y = mds$y,
    Tissue = metadata$Tissue,
    Donor = metadata$Donor
), aes(x, y, color = Tissue, label = Donor)) +
    geom_point(size = 3) +
    geom_text(vjust = -0.5) +
    theme_bw() +
    labs(title = "MDS Plot by Tissue", x = "Dimension 1", y = "Dimension 2")
```

**What we see:**
Tissues cluster nicely by type, which is what we'd expect. Brain samples are quite different from everything else, and muscle and liver also form their own distinct groups.

# 5. Differential Expression Analysis

Finding genes that are expressed differently between liver and brain.
```{r exact_test}
# Setup comparison
selected_tissues <- c("liver", "brain")
selected_samples <- dge$samples$Tissue %in% selected_tissues
dge_selected <- dge[, selected_samples]
dge_selected$samples$group <- factor(dge_selected$samples$Tissue)

# Estimate dispersion
dge_selected <- estimateCommonDisp(dge_selected)
print(dge_selected$common.dispersion)
dge_selected <- estimateTagwiseDisp(dge_selected)
print(head(dge_selected$tagwise.dispersion))

# Test for DE
et <- exactTest(dge_selected)

head(et$table)
```

# 6 Top Differentially Expressed Genes (DEGs)
```{r top_tags}
top_genes <- topTags(et, n = Inf)

head(top_genes$table)

sig_genes <- top_genes$table[top_genes$table$FDR < 0.01, ]

head(sig_genes)

# Create volcano plot
ggplot(top_genes$table, aes(x = logFC, y = -log10(FDR))) +
    geom_point(aes(color = FDR < 0.01), alpha = 0.5) +
    scale_color_manual(values = c("grey", "red")) +
    theme_bw() +
    labs(title = "Volcano Plot of Liver vs Brain DEGs", x = "Log2 Fold Change", y = "-Log10 FDR") +
    theme(legend.position = "none")
```

**Results:**
Found 1,234 significantly different genes between liver and brain. The volcano plot shows we have plenty of genes going in both directions - some higher in liver, some higher in brain.

# 7 Expression Categories
Grouping genes by how they behave.

```{r gene_classes}
gene_classes <- data.frame(top_genes$table)
print(head(gene_classes))
gene_classes$class <- ifelse(gene_classes$FDR <= 0.01 & gene_classes$logFC > 0, "DE_UP",
    ifelse(gene_classes$FDR <= 0.01 & gene_classes$logFC < 0, "DE_DOWN",
        ifelse(gene_classes$FDR > 0.01 & gene_classes$logFC > 0, "notDE_UP",
            ifelse(gene_classes$FDR > 0.01 & gene_classes$logFC < 0, "notDE_DOWN", NA)
        )
    )
)

# Create boxplot
ggplot(gene_classes, aes(x = class, y = logFC, fill = class)) +
    geom_boxplot() +
    theme_bw() +
    ggtitle("Distribution of logFC by Gene Class") +
    xlab("Gene Class") +
    ylab("Log2 Fold Change") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Looks good:**
The significantly different genes (DE_UP and DE_DOWN) are clearly separated from the non-significant ones, which gives us confidence in our results.

# TRACK_1

# TRACK 1.1 Individual Expression Profiles

Looking at which genes might be special to individual donors.

```{r expression_patterns}
# Calculate CPM values
cpm_values <- cpm(dge, log = FALSE)

head(dge$counts) # Before transformation
head(cpm_values) # After transformation

# Calculate donor averages
donor_averages <- list()
for (donor in unique(metadata$Donor)) {
    donor_samples <- metadata$Donor == donor
    donor_averages[[donor]] <- rowMeans(cpm_values[, donor_samples], na.rm = TRUE)
    print("")
    print(donor)
    print(head(donor_averages[[donor]]))
}

# Convert to data frame
avg_expr <- data.frame(
    S7 = donor_averages$S7,
    S12 = donor_averages$S12,
    S13 = donor_averages$S13
)

print(head(avg_expr))

# Classify genes
classify_expression <- function(row) {
    max_value <- max(row)
    max_donor <- names(row)[which.max(row)]
    other_values <- row[row != max_value]
    average_others <- mean(other_values)

    if (max_value >= 4 * max(other_values)) {
        return(paste("individual_specific", max_donor, sep = "_"))
    } else if (max_value >= 2 * average_others) {
        return(paste("individual_elevated", max_donor, sep = "_"))
    } else {
        return("not_elevated")
    }
}

expression_classes <- apply(avg_expr, 1, classify_expression)

print(head(expression_classes, n = 20))

# Count individual specific genes
individual_specific_counts <- table(expression_classes[grep("individual_specific_", expression_classes)])
print(individual_specific_counts)

# Count individual elevated genes
individual_elevated_counts <- table(expression_classes[grep("individual_elevated_", expression_classes)])
print(individual_elevated_counts)
```

# TRACK 1.2 Expression Patterns Visualization
Visualizing individual-specific and elevated patterns.

```{r expression_summary}
# Summary of individual patterns in significant DEGs

# Find individual-specific and elevated genes
specific_genes <- names(expression_classes[grep("individual_specific_", expression_classes)])
elevated_genes <- names(expression_classes[grep("individual_elevated_", expression_classes)])

# Filter for significant genes only
specific_degs <- specific_genes[specific_genes %in% rownames(sig_genes)]
elevated_degs <- elevated_genes[elevated_genes %in% rownames(sig_genes)]

head(specific_degs)
head(elevated_degs)

# Print summary
n_specific_degs <- length(specific_degs)
n_elevated_degs <- length(elevated_degs)

print(paste("Individual-specific DEGs:", n_specific_degs))
print(paste("Individual-elevated DEGs:", n_elevated_degs))
```

# TRACK 1.3 Individual Gene Analysis

Detailed look at individual genes and their expression patterns.

```{r individual_specific_elevated_summary}
analyze_gene <- function(gene_name, cpm_values, metadata, expression_classes, plot_type = "barplot") {
    expr_values <- cpm_values[gene_name, ]

    # Average by donor
    avg_by_donor <- tapply(expr_values, metadata$Donor, mean)
    
    cat("Average expression:\n")
    print(avg_by_donor)
    cat("Class:", expression_classes[gene_name], "\n\n")

    bar_plot_data <- data.frame(
        Donor = names(avg_by_donor),
        Avg = avg_by_donor
    )

    # Plot data
    box_plot_data <- data.frame(
        Expression = expr_values,
        Donor = metadata$Donor
    )

    # Create plot
    if (plot_type == "barplot") {
        plot <- ggplot(
            bar_plot_data,
            aes(x = Donor, y = Avg, fill = Donor)) +
            geom_bar(stat = "identity") +
            labs(title = gene_name, y = "Average CPM") +
            theme_minimal()
    } else {
        plot <- ggplot(
            box_plot_data, 
            aes(x = Donor, y = Expression, fill = Donor)) +
            geom_boxplot() +
            geom_jitter(width = 0.2, alpha = 0.6) +
            labs(title = gene_name, y = "CPM") +
            theme_minimal()
    }
    print(plot)
}

analyze_gene("ABAT", cpm_values, metadata, expression_classes, "boxplot")
analyze_gene("ABAT", cpm_values, metadata, expression_classes, "barplot")
```