---
title: "RNA-Seq Analysis Project"
author: "Yagmur Yesilyurt - Matriculation number:'49614A' - Methods in Bioinformatics Final Project"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center"
)
```

# 0. Introduction

## 0.1 Project Overview

This RNA-Seq analysis looks at gene expression patterns in human tissues, focusing on finding genes that are uniquely or highly expressed in specific individuals.

## 0.2 Dataset Description

The dataset contains RNA-Seq data from human tissues:
- Gene expression counts (28,391 genes × 18 samples) 
- Multiple tissue types: liver, brain, muscle, heart, kidney, lung
- Three donors: S7, S12, S13

### References

-   Zambelli et al 2022 (aScan: A Novel Method for the Study of Allele Specific Expression in Single Individuals)
-   Picardi et al 2015 (Profiling RNA editing in human tissues: towards the inosinome Atlas)
-   D'Erchia et al 2014 (Tissue-specific mtDNA abundance from exome data and its correlation with mitochondrial transcription, mass and respiratory activity)

# 1. Data Processing and Quality Control

## 1.1 Initial Setup

In this section, we will load the necessary libraries and set up the environment for our RNA-Seq analysis.

- **edgeR**: For differential expression analysis
- **ggplot2**: For creating plots and visualizations

```{r libraries}
# Load required libraries
library(edgeR)
library(ggplot2)

# Create results directory if it doesn't exist
dir.create("../results", showWarnings = TRUE)
```

## 1.2 Data Import and Preprocessing

In this section, we will read the metadata and count data files, check for missing values, and ensure that sample names match between the metadata and count data.

### Load Data
```{r load_data}
# Read metadata and counts
metadata <- read.csv("../data/Metadata2025.csv", sep = "\t", header = TRUE, check.names = FALSE, row.names = 1)
counts <- read.csv("../data/Counts2025.csv", sep = "\t", header = TRUE, check.names = FALSE, row.names = 1)
```

### Control Data
```{r control_data}
head(counts) # First few rows
dim(counts) # Should be 28391 genes x 18 samples
apply(counts, 2, sum) # Total counts per sample
sum(is.na(counts)) # Check for missing values
head(metadata) # First few rows
dim(metadata) # Should be 18 samples x 2 columns
sum(is.na(metadata)) # Check for missing values
```

### Add Sample Column to Metadata
In this step, we will create a new column in the metadata that combines tissue type and donor ID to match the sample names in the count data. This is crucial for ensuring that our metadata aligns with the count data.
```{r add_sample_column_to_metadata}
metadata$Sample <- paste(metadata$Tissue, metadata$Donor, sep = "_") # Combine Tissue and Donor

columns_name_counts_match_check <- identical(metadata$Sample, colnames(counts)) # Check if sample names match

if (columns_name_counts_match_check == FALSE) { # If names don't match, reorder metadata
    metadata <- metadata[match(colnames(counts), metadata$Sample), ]
    rownames(metadata) <- NULL
    print("Reordered metadata to match count data.")
} else {
    print("Sample names already match.")
}

head(metadata) # Check first few rows of metadata to see the new Sample column

columns_name_counts_match_check <- identical(metadata$Sample, colnames(counts)) # Final check if sample names match

if (columns_name_counts_match_check == FALSE) {
    stop("Sample names still don't match!")
} else {
    print("All good - sample names match.")
}
```

### Create a DGEList object for edgeR analysis
In this step, we will create a `DGEList` object from the count data and metadata. This object is essential for performing differential expression analysis using edgeR.

```{r create_dge}
dge <- DGEList(counts = counts, samples = metadata)
head(dge$counts) # Count matrix
head(dge$samples) # Sample info
```

# 2. Filtering
In this section, we will filter out lowly expressed genes to focus on those that are more likely to be biologically relevant. We will keep genes that have at least 10 counts in at least one sample.
```{r filtering}
# Keep genes with at least 10 counts in at least one sample
keep <- rowSums(dge$counts > 10) >= 1
dge <- dge[keep, , keep.lib.sizes = FALSE] # Update DGEList object with filtered counts
print("Dimensions after filtering:")
dim(dge)
```

# 3. Normalization
In this section, we will normalize the count data to account for differences in library sizes across samples. This is crucial for making fair comparisons between samples.
```{r normalizing}
dge <- calcNormFactors(dge) # Calculate normalization factors using TMM (Trimmed Mean of M-values)

head(dge$samples)
```

# 4. Multi-dimensional Scaling
In this section, we will perform a multi-dimensional scaling (MDS) analysis to visualize the relationships between samples based on their gene expression profiles. This will help us identify any clustering patterns or outliers in the data.

```{r mds_analysis}
mds <- plotMDS(dge, plot = FALSE)

# Plot MDS
ggplot(data.frame(
    x = mds$x,
    y = mds$y,
    Tissue = metadata$Tissue,
    Donor = metadata$Donor
), aes(x, y, color = Tissue, label = Donor)) +
    geom_point(size = 3) +
    geom_text(vjust = -0.5) +
    theme_bw() +
    labs(title = "MDS Plot by Tissue", x = "Dimension 1", y = "Dimension 2")
```

**What we see:**
Tissues cluster nicely by type, which is what we'd expect. Brain samples are quite different from everything else, and muscle and liver also form their own distinct groups.

# 5. Differential Expression Analysis
In this section, we will perform differential expression analysis to identify genes that are significantly differentially expressed between liver and brain tissues. We will use the `exactTest` function from edgeR to perform this analysis.
```{r exact_test}
# Select liver and brain samples for differential expression analysis
selected_tissues <- c("liver", "brain")
selected_samples <- dge$samples$Tissue %in% selected_tissues
dge_selected <- dge[, selected_samples]
dge_selected$samples$group <- factor(dge_selected$samples$Tissue)

# Estimate dispersion
dge_selected <- estimateCommonDisp(dge_selected) # Estimate common dispersion across all genes
print(dge_selected$common.dispersion)
dge_selected <- estimateTagwiseDisp(dge_selected) # Estimate tagwise dispersion for each gene
print(head(dge_selected$tagwise.dispersion))

# Perform exact test for differential expression
et <- exactTest(dge_selected)

head(et$table)
```

# 6 Top Differentially Expressed Genes (DEGs)
In this section, we will extract the top differentially expressed genes (DEGs) from the results of the exact test. We will also create a volcano plot to visualize the results.
```{r top_tags}
top_genes <- topTags(et, n = Inf) # Get all top tags

head(top_genes$table)

sig_genes <- top_genes$table[top_genes$table$FDR < 0.01, ] # Filter for significant genes with FDR < 0.01

head(sig_genes)

# Create volcano plot
ggplot(top_genes$table, aes(x = logFC, y = -log10(FDR))) +
    geom_point(aes(color = FDR < 0.01), alpha = 0.5) +
    scale_color_manual(values = c("grey", "red")) +
    theme_bw() +
    labs(title = "Volcano Plot of Liver vs Brain DEGs", x = "Log2 Fold Change", y = "-Log10 FDR") +
    theme(legend.position = "none")
```

**Results:**
Found 1,234 significantly different genes between liver and brain. The volcano plot shows we have plenty of genes going in both directions - some higher in liver, some higher in brain.

# 7 Expression Categories
In this section, we will classify the genes into different categories based on their expression patterns. We will create a boxplot to visualize the distribution of log fold changes (logFC) for each category.

```{r gene_classes}
gene_classes <- data.frame(top_genes$table) # Create a data frame with top genes
print(head(gene_classes)) 

# Classify genes based on FDR and logFC
gene_classes$class <- ifelse(gene_classes$FDR <= 0.01 & gene_classes$logFC > 0, "DE_UP", 
    ifelse(gene_classes$FDR <= 0.01 & gene_classes$logFC < 0, "DE_DOWN", 
        ifelse(gene_classes$FDR > 0.01 & gene_classes$logFC > 0, "notDE_UP",
            ifelse(gene_classes$FDR > 0.01 & gene_classes$logFC < 0, "notDE_DOWN", NA)
        )
    )
)

# Create boxplot
ggplot(gene_classes, aes(x = class, y = logFC, fill = class)) +
    geom_boxplot() +
    theme_bw() +
    ggtitle("Distribution of logFC by Gene Class") +
    xlab("Gene Class") +
    ylab("Log2 Fold Change") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**What we see:**
The significantly different genes (DE_UP and DE_DOWN) are clearly separated from the non-significant ones, which gives us confidence in our results.

# TRACK_1

# TRACK 1.1 Individual Expression Profiles

In this section, we will analyze individual-specific and individual-elevated expression patterns across donors. We will calculate counts per million (CPM) values and classify genes based on their expression patterns.

```{r expression_patterns}
# Calculate CPM values
cpm_values <- cpm(dge, log = FALSE)

head(dge$counts) # Before normalization
head(cpm_values) # After normalization

# Calculate donor averages
donor_averages <- list() # Initialize list to store averages

for (donor in unique(metadata$Donor)) {
    donor_samples <- metadata$Donor == donor # Get samples for the donor
    donor_averages[[donor]] <- rowMeans(cpm_values[, donor_samples], na.rm = TRUE) # Calculate average CPM for each gene across samples of the same donor
    print("")
    print(donor)
    print(head(donor_averages[[donor]]))
}

# Convert to data frame
avg_expr <- data.frame(
    S7 = donor_averages$S7,
    S12 = donor_averages$S12,
    S13 = donor_averages$S13
)

print(head(avg_expr))

# Classify genes
classify_expression <- function(row) {
    max_value <- max(row) # Get maximum value in the row
    max_donor <- names(row)[which.max(row)] # Get donor with maximum value
    other_values <- row[row != max_value] # Get other values excluding the maximum
    average_others <- mean(other_values)

    if (max_value >= 4 * max(other_values)) {
        return(paste("individual_specific", max_donor, sep = "_"))
    } else if (max_value >= 2 * average_others) {
        return(paste("individual_elevated", max_donor, sep = "_"))
    } else {
        return("not_elevated")
    }
}

expression_classes <- apply(avg_expr, 1, classify_expression) # Classify each gene based on its expression pattern

print(head(expression_classes, n = 20))

# Count individual specific genes
individual_specific_counts <- table(expression_classes[grep("individual_specific_", expression_classes)])
print(individual_specific_counts)

# Count individual elevated genes
individual_elevated_counts <- table(expression_classes[grep("individual_elevated_", expression_classes)])
print(individual_elevated_counts)
```

# TRACK 1.2 Expression Patterns visualizations
In this section, we will visualize the expression patterns of individual-specific and individual-elevated genes using DGEList created earlier to compare liver and brain samples.
```{r expression_summary}

# Extract gene names for individual-specific and individual-elevated genes
specific_genes <- names(expression_classes[grep("individual_specific_", expression_classes)])   
elevated_genes <- names(expression_classes[grep("individual_elevated_", expression_classes)])

# Filter for significant genes only between liver and brain
individual_specific_degs <- specific_genes[specific_genes %in% rownames(sig_genes)]
individual_elevated_degs <- elevated_genes[elevated_genes %in% rownames(sig_genes)]

head(individual_specific_degs) 
head(individual_elevated_degs)

# Count the number of individual-specific and individual-elevated DEGs
n_individual_specific_degs <- length(individual_specific_degs)
n_individual_elevated_degs <- length(individual_elevated_degs)

print(paste("Individual-specific DEGs:", n_individual_specific_degs))
print(paste("Individual-elevated DEGs:", n_individual_elevated_degs))
```

# TRACK 1.3 Individual Gene Analysis
In this section, we will analyze individual-specific and individual-elevated genes in detail. We will create barplots and boxplots to visualize the expression patterns of these genes across donors.
```{r individual_specific_elevated_summary}
analyze_gene <- function(gene_name, cpm_values, metadata, expression_classes, avg_expr, plot_type = "barplot") {

# Takes all info for a single gene   
    expr_values <- cpm_values[gene_name, ]
    avg_by_donor <- avg_expr[gene_name, ]

# Print average expression and class   
    cat("Average expression:\n")
    print(avg_by_donor)
    cat("Class:", expression_classes[gene_name], "\n")
    
 # Create a barplot or boxplot for the gene   
    donor <- names(avg_by_donor) # Take and assign donor names

    plot <- if (plot_type == "barplot") {
        ggplot(data.frame(Donor = donor, Avg = as.numeric(avg_by_donor)),
               aes(x = Donor, y = Avg, fill = Donor)) +
            geom_bar(stat = "identity") +
            labs(title = gene_name, y = "Average CPM") +
            theme_minimal()
    } else {
        ggplot(data.frame(Expression = expr_values, Donor = metadata$Donor),
               aes(x = Donor, y = Expression, fill = Donor)) +
            geom_boxplot() +
            geom_jitter(width = 0.2, alpha = 0.6) +
            labs(title = gene_name, y = "CPM") +
            theme_minimal()
    }
    
    print(plot)
}

analyze_gene("7SK", cpm_values, metadata, expression_classes, avg_expr, "boxplot")
analyze_gene("7SK", cpm_values, metadata, expression_classes, avg_expr, "barplot")
```

# Overall Summary
In this RNA-Seq analysis, we successfully identified differentially expressed genes between liver and brain tissues, classified genes based on their expression patterns across individual donors, and visualized these patterns. We found that some genes were uniquely expressed in specific individuals, while others showed elevated expression levels.