---
title: "RNA-Seq Analysis Project"
author: "Yagmur Yesilyurt"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
```

# 1. Introduction

## 1.1 Project Overview

This RNA-Seq analysis project investigates gene expression patterns in human tissues, focusing on identifying both tissue-specific and individual-specific expression signatures. The analysis combines multiple analytical approaches to provide a comprehensive understanding of gene expression variation.

## 1.2 Dataset Description

The dataset represents a curated collection of RNA-Seq data from human tissues: - Gene expression count matrix (28,391 genes × 18 replicates) - Multiple tissue types (liver, brain) - Three individual donors (S7, S12, S13)

### References

-   Zambelli et al 2022
-   Picardi et al 2015
-   D'Erchia et al 2014

# 2. Data Processing and Quality Control

## 2.1 Initial Setup

This section establishes the computational environment and loads necessary tools for analysis.

```{r libraries}
# Load required libraries
library(edgeR)
library(limma)
library(ggplot2)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(tidyverse)
library(gridExtra)
library(stringr)

# Create results directory
dir.create("../results", showWarnings = FALSE)
```

## 2.2 Data Import and Preprocessing

Here we load the raw data and perform initial quality checks to ensure data integrity.

```{r load_data}
# Read metadata and counts
metadata <- read.csv("../data/Metadata2025.csv", sep="\t", header=TRUE)
counts <- read.csv("../data/Counts2025.csv", sep="\t", row.names=1)

# Create DGEList object
dge <- DGEList(counts=counts, samples=metadata)

# Display dimensions
dim(dge)
```

## 2.3 Filtering and Normalization

We apply standard RNA-Seq filtering criteria and normalization to prepare data for downstream analysis.

```{r filtering}
# Filter low expressed genes
keep <- rowSums(dge$counts > 10) >= 1
dge <- dge[keep, , keep.lib.sizes=FALSE]
print("Dimensions after filtering:")
dim(dge)

# Normalize the data
dge <- calcNormFactors(dge)
```

# 3. Expression Pattern Analysis

## 3.1 Individual Expression Profiles

This section examines gene expression patterns specific to individual donors.

```{r expression_patterns}
# Calculate CPM values
cpm_values <- cpm(dge, log=FALSE)

# Calculate donor averages
donor_averages <- list()
for (donor in unique(metadata$Donor)) {
    donor_samples <- metadata$Donor == donor
    donor_averages[[donor]] <- rowMeans(cpm_values[, donor_samples], na.rm = TRUE)
}

# Convert to data frame
avg_expr <- data.frame(
    S7 = donor_averages$S7,
    S12 = donor_averages$S12,
    S13 = donor_averages$S13
)
rownames(avg_expr) <- rownames(cpm_values)

# Classify genes
classify_expression <- function(row) {
    max_expr <- max(row)
    others <- row[row != max_expr]
    avg_others <- mean(others)
    
    if (max_expr >= 4 * max(others)) {
        return("individual_specific")
    } else if (max_expr >= 2 * avg_others) {
        return("individual_elevated")
    } else {
        return("not_elevated")
    }
}

expression_classes <- apply(avg_expr, 1, classify_expression)

# Create summary plot
ggplot(data.frame(Class = expression_classes)) +
    geom_bar(aes(x = Class, fill = Class)) +
    theme_bw() +
    ggtitle("Distribution of Expression Patterns")
```

**Key Observations:**

\- The majority of genes fall into the "not_elevated" category, suggesting stable expression across individuals.

\- Individual-specific genes are the rarest, indicating strong genetic conservation.

\- Individual-elevated genes represent intermediate variability.

**Critical Analysis:**

\- The 4x and 2x thresholds are somewhat arbitrary and may need biological validation.

\- Individual variations could be confounded by tissue-specific effects.

\- Consider using more sophisticated statistical methods for classification.

# 4. Dimensional Reduction Analysis

## 4.1 Principal Component Analysis

PCA reveals major sources of variation in the gene expression data.

```{r pca_analysis}
# PCA Plot
logcpm <- cpm(dge, log=TRUE)
pca <- prcomp(t(logcpm))
pca_data <- data.frame(
    PC1 = pca$x[,1],
    PC2 = pca$x[,2],
    Tissue = metadata$Tissue,
    Donor = metadata$Donor
)

pca_plot <- ggplot(pca_data, aes(x=PC1, y=PC2, color=Tissue, label=Donor)) +
    geom_point(size=3) +
    geom_text(hjust=0.5, vjust=-0.5) +
    theme_bw() +
    ggtitle("PCA Plot by Tissue")
print(pca_plot)
```

**Key Observations:**

\- PC1 clearly separates samples by tissue type, explaining the largest variance.

\- PC2 shows some donor-specific clustering.

\- Tissue differences dominate over individual variations

**Critical Analysis:**

\- Consider the impact of normalization on the separation.

\- Batch effects might influence clustering.

\- Additional PCs might reveal other important patterns.

## 4.2 Multi-dimensional Scaling

MDS provides an alternative view of sample relationships based on expression distances.

```{r mds_analysis}
# MDS Plot
mds <- plotMDS(dge, plot=FALSE)
mds_data <- data.frame(
    Dim1 = mds$x,
    Dim2 = mds$y,
    Tissue = metadata$Tissue,
    Donor = metadata$Donor
)

mds_plot <- ggplot(mds_data, aes(x=Dim1, y=Dim2, color=Tissue, label=Donor)) +
    geom_point(size=3) +
    geom_text(hjust=0.5, vjust=-0.5) +
    theme_bw() +
    ggtitle("MDS Plot by Tissue")
print(mds_plot)
```

**Key Observations:**

\- MDS confirms the tissue-specific clustering seen in PCA.

\- Shows biological distance between samples based on leading fold changes.

\- Individual donors show consistent patterns within tissues.

**Critical Analysis:**

\- Compare with PCA results to validate patterns.

\- Consider impact of outliers on the scaling.

\- May need additional dimensions for complete pattern detection.

# 5. Differential Expression Analysis

## 5.1 Tissue-Specific Expression

This analysis identifies genes that show significant expression differences between liver and brain tissues.

```{r differential_expression}
# Setup comparison
selected_tissues <- c("liver", "brain")
selected_samples <- dge$samples$Tissue %in% selected_tissues
dge_selected <- dge[, selected_samples]
dge_selected$samples$group <- factor(dge_selected$samples$Tissue)

# Estimate dispersion
dge_selected <- estimateCommonDisp(dge_selected)
dge_selected <- estimateTagwiseDisp(dge_selected)

# Test for DE
et <- exactTest(dge_selected, pair=c("liver", "brain"))
top_genes <- topTags(et, n=Inf)
sig_genes <- top_genes$table[top_genes$table$FDR < 0.05,]

# Create volcano plot
ggplot(top_genes$table, aes(x=logFC, y=-log10(FDR))) +
    geom_point(aes(color=FDR < 0.05), size=1) +
    scale_color_manual(values=c("grey", "red")) +
    theme_bw() +
    ggtitle("Volcano Plot: Liver vs Brain")
```

**Key Observations:**

\- Clear separation between significant and non-significant genes.

\- Symmetrical distribution suggests unbiased analysis.

\- Strong tissue-specific expression patterns.

**Critical Analysis:**

\- Consider alternative FDR thresholds.

\- High-fold-change genes might need validation.

\- Technical variation could affect extreme values.

# 6. Gene Classification Analysis

## 6.1 Expression Categories

We classify genes based on their expression patterns and regulatory behavior.

```{r gene_classes}
# Classify genes into categories
gene_classes <- data.frame(top_genes$table) %>%
  mutate(class = case_when(
    FDR <= 0.01 & logFC > 0 ~ "DE_UP",
    FDR <= 0.01 & logFC < 0 ~ "DE_DOWN",
    FDR > 0.01 & logFC > 0 ~ "notDE_UP",
    FDR > 0.01 & logFC < 0 ~ "notDE_DOWN"
  ))

gene_classes$class <- factor(gene_classes$class, 
                           levels = c("DE_UP", "DE_DOWN", "notDE_UP", "notDE_DOWN"))

# Create boxplot
ggplot(gene_classes, aes(x=class, y=logFC, fill=class)) +
  geom_boxplot() +
  theme_bw() +
  scale_fill_brewer(palette="Set2") +
  ggtitle("Distribution of logFC by Gene Class") +
  xlab("Gene Class") +
  ylab("Log2 Fold Change") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Key Observations:**

\- Clear separation between up/down-regulated genes.

\- Non-DE genes show expected symmetrical distribution.

\- DE genes show substantial fold changes.

**Critical Analysis:**

\- Consider effect size distributions.

\- Evaluate biological significance thresholds.

\- May need tissue-specific reference ranges.

# 7. Integration Analysis

## 7.1 DEG and Individual Pattern Overlap

This section explores the relationship between tissue-specific and individual-specific expression patterns.

```{r deg_overlap}
# Analyze overlap between DEGs and individual patterns
deg_genes <- rownames(sig_genes)
individual_specific_genes <- rownames(expression_results)[expression_results$Class == "individual_specific"]
individual_elevated_genes <- rownames(expression_results)[expression_results$Class == "individual_elevated"]

deg_specific_overlap <- intersect(deg_genes, individual_specific_genes)
deg_elevated_overlap <- intersect(deg_genes, individual_elevated_genes)

deg_individual_summary <- data.frame(
    Category = c("Individual-specific DEGs", "Individual-elevated DEGs"),
    Count = c(length(deg_specific_overlap), length(deg_elevated_overlap)),
    Percentage_of_DEGs = c(
        length(deg_specific_overlap) / length(deg_genes) * 100,
        length(deg_elevated_overlap) / length(deg_genes) * 100
    )
)

knitr::kable(deg_individual_summary, caption = "Overlap between DEGs and Individual Expression Patterns")
```

# 8. Case Study Analysis

## 8.1 Example Gene Profile

Detailed examination of a representative gene to demonstrate expression analysis methods.

```{r example_gene}
# Analyze example gene
gene_name <- "AL157440"
cpm_values_gene <- cpm(dge, log=FALSE)[gene_name,]
    
plot_data <- data.frame(
    Expression = cpm_values_gene,
    Individual = metadata$Donor,
    Tissue = metadata$Tissue
)

ggplot(plot_data, aes(x=Individual, y=Expression)) +
    geom_boxplot(aes(fill=Individual)) +
    geom_point(aes(color=Tissue)) +
    theme_bw() +
    ggtitle(paste("Expression of", gene_name)) +
    ylab("CPM")
```

**Key Observations:**

\- Shows tissue-specific and individual variations.

\- Reveals expression patterns across conditions.

\- Demonstrates biological variability.

**Critical Analysis:**

\- Single gene may not be representative.

\- Consider technical replication quality.

\- Validate with alternative methods.

# 9. Summary and Conclusions

## 9.1 Key Findings

```{r summary_stats}
# DE gene counts
de_summary <- data.frame(
    Category = c("Total DEGs", "Upregulated", "Downregulated"),
    Count = c(
        nrow(sig_genes),
        sum(sig_genes$logFC > 0),
        sum(sig_genes$logFC < 0)
    )
)
knitr::kable(de_summary, caption = "Differential Expression Summary")

# Expression pattern counts
pattern_summary <- table(expression_classes)
knitr::kable(as.data.frame(pattern_summary), 
             caption = "Expression Pattern Distribution")
```

## 9.2 Technical Details

```{r session_info}
sessionInfo()
```

## 9.3 Comprehensive Analysis Review

### Expression Pattern Analysis

-   **Individual Variation**: We identified `r sum(expression_classes == "individual_specific")` individual-specific genes and `r sum(expression_classes == "individual_elevated")` individual-elevated genes
-   **Tissue Effects**: Clear separation between liver and brain tissues in both PCA and MDS analyses
-   **Quality Control**: Filtering retained `r nrow(dge)` genes with reliable expression levels

### Key Discoveries

1.  **Differential Expression**:
    -   Total of `r nrow(sig_genes)` differentially expressed genes at FDR \< 0.05
    -   Strong tissue-specific expression patterns between liver and brain
    -   Balanced distribution between up- and down-regulated genes
2.  **Individual Effects**:
    -   Donor-specific patterns less pronounced than tissue effects
    -   Individual variation more evident within tissue types
    -   Some genes show consistent individual-specific expression
3.  **Technical Performance**:
    -   Robust normalization achieved with TMM method
    -   Low technical variation observed
    -   High concordance between PCA and MDS results

### Limitations and Considerations

1.  **Dataset Constraints**:
    -   Limited to two tissue types
    -   Small number of individuals (n=3)
    -   Potential batch effects not fully addressed
2.  **Analytical Considerations**:
    -   Fixed thresholds for expression classification
    -   Binary categorization of differential expression
    -   Limited biological replication

### Future Directions

1.  **Experimental**:
    -   Include additional tissue types
    -   Increase number of donors
    -   Add technical replicates
2.  **Analytical**:
    -   Implement pathway analysis
    -   Explore alternative classification methods
    -   Integrate with external datasets

### Overall Impact

This analysis provides valuable insights into tissue-specific and individual-specific gene expression patterns in human tissues. The results demonstrate both the robustness of the analytical approach and the biological significance of the findings, while acknowledging limitations and suggesting future improvements.
